Ancestry inference and grouping\\from principal component analysis of genetic data

### 2020-10-18 paper/paper-ancestry-matching.tex
save

\abstract{
	First, we show how Euclidean distance in the Principal Component Analysis (PCA) space is related to $F_{ST}$ between populations.
	Then, we project the genetic data of individuals from the UK Biobank and the POPRES datasets onto the reference populations of the 1000 Genomes Project and show how to accurately infer global ancestry and group individuals in homogeneous clusters. 
	Inferring ancestry for nearly 500,000 individuals is performed in 20 minutes only.
	The proposed method is particularly important for methods that needs to be applied in homogeneous samples.
	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
}

@@

Here, we first compare several distance metrics with the popular $F_{ST}$ statistic between populations.
We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
We do not provide a method to define admixture coefficients nor infer local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of ancestry homogeneous.
This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.

@@

We first compare four measures of genetic dissimilarity using populations of the 1000 Genomes Project (1000G$^\dag$, \cite{10002015global}). 
The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}.
The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shorter distance between pairs of PC scores from the two populations.
The distance between population centers seems to be an appropriate PCA-based distance as this (squared) distance is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:heatmap1} and \ref{fig:heatmap3}). For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.

@@

We project the dataset of interest onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
We recall that this uses a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}. 
Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the distance to these reference population centers (geometric medians$^\dag$). 
Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002. This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest population.

We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
These individuals seem to originate from all over the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
Based on the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}). This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries. 
Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian` ancestries.
Only 47 individuals were wrongly classified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}). 
However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed on the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is wrong (Figure \ref{fig:mismatch}).

We also test the approach proposed in \cite{zhang2020fast} which consists of finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals. When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
Less than 0.5\% could not be matched by their method (Table \ref{tab:ancestry-pred-kNN}).
Of note, they could match much more admixed individuals, whereas they set a high probably aiming at discarding such admixed individuals. 
Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
Finally, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-fine-pred}).
We also applied our ancestry detection technique to the European individuals of the POPRES data \cite[]{nelson2008population}. 
Only 16 out of the 1385 individuals (1.2\%) could not be matched, of which 11 were from East or South-East Europe (Table \ref{tab:ancestry-pred-popres}).
Note that all individuals that we could match were identified as of European ancestry. 
We could also identify accurately sub-regions of Europe; e.g.\ 261 out of 264 Spanish and Portugese individuals were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).

@@

Finally, we show 3 possible ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
The first approach is to simply match individuals that are closed enough to one or more of the 1000G populations, as described in this paper.
We can e.g.\ define a British ancestry cluster by identifying individuals close enough to the GBR 1000G population (Figure \ref{fig:hist1}).
Likewise, we can define ``AFR'', ``AMR'', ``EAS'' and ``SAS'' clusters of individuals who are close enough to any of the corresponding 1000G populations.
We can then verify that these clusters define distinct sets of homogeneous individuals (Figure \ref{fig:grouping}).
Another solution, which does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead, would consist in using reported ancestries and internal PC scores.
Then, a British ancestry cluster could be defined by individuals close enough to the center of the people reported as ``British''  (Figure \ref{fig:hist2}).
We can also define an ``AFR2'' cluster if close enough to the center of either ``Caribbean'', ``African'' or ``Other Black'' individuals, ``EAS2'' if close to the ``Chinese'' cluster, and ``SAS2'' if close to either ``Indian'', ``Pakistani'' or ``Bangladeschi'' clusters.
This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not really ``British'' as self-reported. 
A third solution, useful when we know that the data is composed of a predominant ancestry is to keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
We recall the we are using the geometric median as such center, which is robust to outliers.
Overall, all these techniques approximately identify the same amount of people (Table \ref{tab:ancestry-groups}).

@@

@@

This research has been conducted using the UK Biobank Resource under Application Number 41181.

@@

\subsection*{Location of 1000G populations}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{map-1000G}}
	\caption{Locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population} and represented using R package leaflet. 
	Note that Gujarati Indian from Houston, Texas were manually moved to Gujarat (22.309425, 72.136230), Sri Lankan Tamil from the UK to Sri Lanka (6.927079,	79.861244), and Indian Telugu from the UK to (16.5, 79.5) to better reflect the location of their ancestors.
	Also note that Utah Residents with Northern and Western European Ancestry, Americans of African Ancestry in SW USA, African Caribbeans in Barbados and Mexican Ancestry from Los Angeles USA are not located at their ancestral location. \label{fig:map}}
\end{figure}

@@

@@

\begin{figure}[h]
\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data. The threshold in red assumes an $F_{ST}$ of 0.002. \label{fig:hist1}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals. The threshold in red assumes an $F_{ST}$ of 0.002. \label{fig:hist2}}
\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
	\caption{Histogram of (log) squared distances from the UK Biobank PC scores to the geometric median of the all UKBB individuals. Here we use a threshold at 7, based on visual inspection. \label{fig:hist3}}
\end{figure}

>>>>

\abstract{
	First, we show how Euclidean distance in the Principal Component Analysis (PCA) space is related to $F_{ST}$ between populations.
	Then, we project the genetic data of individuals from the UK Biobank and the POPRES datasets onto the reference populations of the 1000 Genomes Project and show how to accurately infer global ancestry and group individuals in homogeneous clusters. 
	Inferring ancestry for nearly 500,000 individuals is performed in 20 minutes only.
	The proposed method is particularly important for methods that need to be applied in homogeneous samples.
	Finally, we conclude that our method and the community would benefit from having an easy access to a reference dataset with an even better coverage of the worldwide genetic diversity than the 1000 Genomes Project.
}

@@

Here, we first compare several distance metrics with the popular $F_{ST}$ statistic between populations.
We show that the simple Euclidean distance on PC scores is the most appropriate distance to use, then we show how to use it to infer global ancestry and to group individuals in homogeneous sub-populations.
We do not provide a method to infer admixture coefficients nor local ancestry, which are different problems for which there are several existing methods \cite[]{alexander2009fast,frichot2014fast,raj2014faststructure,padhukasahasram2014inferring}.
However, inferring global ancestry in non-admixed individuals is still a very important problem since there are methods that specifically need to be applied in samples of homogeneous ancestry.
This is the case e.g.\ for polygenic score methods that have been shown to underperform when applied to populations not homogeneous to the one used for training \cite[]{martin2017human}.

@@

We first compare four measures of genetic dissimilarity using populations of the 1000 Genomes Project (1000G$^\dag$, \cite{10002015global}). 
The $F_{ST}$$^\dag$ is an ubiquitous measure of genetic dissimilarity between populations and the first measure we use in this comparison.
We report $F_{ST}$ between the 26 1000G populations in tables \ref{tab:fst-AFR}-\ref{tab:fst-SAS}, and the clustering of these populations based on $F_{ST}$ in figure \ref{fig:heatmap0}.
The other three measures compared are distances applied to the PC scores$^\dag$ of the genetic data: 1) the Bhattacharyya distance$^\dag$; 2) the distance between the centers (geometric median$^\dag$) of the two populations; 3) the shortest distance between pairs of PC scores from the two populations.
The distance between population centers appears to be an appropriate PCA-based distance as this distance (squared) is approximately proportional to the $F_{ST}$ (Figure \ref{fig:compare-dist0}) and provides an appropriate clustering of populations (Figure \ref{fig:heatmap2}).
In contrast, the two other Bhattacharyya and shortest distances do not provide as satisfactory results (Figures \ref{fig:compare-dist}, \ref{fig:heatmap1} and \ref{fig:heatmap3}). For example, African Caribbeans in Barbados (ACB) and Americans of African Ancestry in SW USA (ASW) are close to Northern European populations (FIN, CEU and GBR) when using the Bhattacharyya distance.
Using the shortest distance between pairs of individuals in two different populations is very sensitive to outliers being far from the other individuals in the population.
Note that we used 19 PCs for computing all PCA-based distances based on visual inspection of PC scores, as recommended in \cite{prive2020efficient}. 
With 2 to 4 PCs, we are able to adequately separate distant populations, but not the closest ones. For example, when using 4 PCs, there are pairs of populations with an $F_{ST}$ of \textasciitilde0.02 while their PC centers are at the same position (Figure \ref{fig:compare-distK}).

\begin{figure}[h]
	\centerline{\includegraphics[width=0.7\textwidth]{compare-Euclidean-to-Fst}}
	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. \label{fig:compare-dist0}}
\end{figure}

@@

We project the dataset of interest onto the PCA space of the 1000G data using the fast tools developed in \cite{prive2020efficient}.
We recall that this uses a correction for shrinkage in projected PC scores, which has been shown to be particularly important when using PC scores for ancestry estimation \cite[]{zhang2020fast}. 
Based on the results of the previous section, we propose to assign individual ancestry to one of the 26 1000G populations based on the distance to these reference population centers (geometric medians$^\dag$). 
Since we showed previously that (squared) distances in the PCA space are proportional to $F_{ST}$, we can set a threshold on these distances that would correspond approximately to an $F_{ST}$ of e.g.\ 0.002. This threshold is close to the dissimilarity between Spanish and Italian people ($F_{ST}$(IBS, TSI) of 0.0015).
When an individual is not close enough to any of the 26 1000G populations, we leave its ancestry inference as unknown, otherwise we assign this individual to the closest reference population center.

We first perform ancestry estimation for the individuals in the UK Biobank$^\dag$.
These individuals seem to originate from all over the world when we project onto the PCA space of the 1000G (Figure \ref{fig:proj-UKBB}). 
Self-reported ancestry (Field 21000) is available for almost all individuals, with only 1.6\% with unknown or mixed ancestry.
Based on the threshold defined before, we could not infer ancestry for 4.6\% of all 488,371 individuals.
More precisely, among ``British'', ``Irish'' and ``White'' ancestries, this represented respectively 2.2\%, 3.3\% and 7.9\% (Tables \ref{tab:infer-UKBB-superpop} and \ref{tab:ancestry-fine-pred}). This also represented 3.3\% for ``Chinese'', 13.8\% for ``Indian'' and 17.8\% for ``African'' ancestries. 
Finally, mixed ancestries were particularly difficult to match to any of the 1000G populations, e.g.\ 97.3\% unmatched within `White and Black African` and 93.0\% within `White and Asian` ancestries.
Only 47 individuals were misclassified in ``super'' population of the 1000G; e.g.\ six ``British'' were classified as South Asians, one ``Chinese'' as European and 25 ``Caribbean'' as South Asian by our method (Table \ref{tab:infer-UKBB-superpop}). 
However, when comparing the location of these mismatched individuals to the rest of individuals on the PCA space computed within the UK Biobank \cite[]{bycroft2017genome}, it seems more probable that our ancestry estimate is exact while the self-reported ancestry is not matching the underlying genetic ancestry for these individuals (Figure \ref{fig:mismatch}).

We also test the approach proposed in \cite{zhang2020fast} which consists in finding the 20 nearest neighbors in 1000G and computing the frequency of (super) population membership, weighted by the inverse distance to these 20 closest 1000G individuals. When this probability is less than 0.875, they leave the ancestry as unknown, aiming at discarding admixed individuals.
Less than 0.5\% could not be matched by their method (Table \ref{tab:ancestry-pred-kNN}).
Of note, they could match much more admixed individuals, whereas they set a high probably aiming at discarding such admixed individuals. 
Morever, there are many more discrepancies between their method and the self-reported ancestry in the UK Biobank (Table \ref{tab:ancestry-pred-kNN}) compared to the previous results with our method (Table  \ref{tab:infer-UKBB-superpop}).
Finally, our method is able to accurately differentiate between sub-continental populations such as differentiating between Pakistani, Bangladeshi and Chinese people (Table \ref{tab:ancestry-fine-pred}).
We also applied our ancestry detection technique to the European individuals of the POPRES data \cite[]{nelson2008population}. 
Only 16 out of the 1385 individuals (1.2\%) could not be matched, of which 11 were from East or South-East Europe (Table \ref{tab:ancestry-pred-popres}).
Note that all individuals that we could match were identified as of European ancestry. 
We could also identify accurately sub-regions of Europe; e.g.\ 261 out of 264 Spanish and Portugese individuals were identified as ``Iberian Population in Spain'' (EUR\_IBS, Table \ref{tab:ancestry-pred-popres}).

@@

Finally, we show several ways how to use our ancestry inference method for grouping genetically homogeneous individuals.
The first approach is to simply match individuals that are close enough to one of the 1000G populations, as described previously.

Alternatively, one could use the internal PC scores and the self-reported ancestries or countries of birth. This solution does not require projecting individuals to the 1000G, but does require computing PC scores within the dataset instead.
We can define centers of the seven self-reported ancestry groups: British, Indian, Pakistani, Bangladeshi, Chinese, Caribbean and African; then match all individuals to one of these centers (or none if an individual is far from all of them).
This enables e.g.\ to capture a larger set of individuals who are close enough to British people (e.g.\ Irish people), while discarding individuals whose genetic ancestry is not matching the self-reported ancestry (TODO: TABLE).
The resulting clusters are presented on the PC space in figure \ref{fig:grouping}.

One could do the same using the countries of birth (Field 20115) instead of the self-reported ancestries.
Again, the country of birth may sometime not reflect the ancestral origin. Therefore, we first compute the robust centers (geometric medians) of all countries with at least 300 individuals. Then, we cluster these countries based on their distance in the PCA space to make sure of their validity as proxies for genetic ancestry and to choose a small subset with good coverage of the overall dissimilarities (TODO FIGURE HEATMAP).
Based on the previous clustering and the available sample sizes, we chosed to use the centers from the following eight countries as reference: the United Kingdom, Poland, Iran, Italy, India, China, ``Caribbean'' and Nigeria.
Interestingly, the individuals in the clusters from Iran and Poland were not previously identified when using the self-reported ancestries, while some individuals previously identified as ``Bangladeshi'' are not matched to India (TODO FIGURES).
2.8\% of individuals could not be matched using countries of birth compared to 3.7\% when using the self-reported ancestry (TODO TABLES).

Finally, when we know that the data is composed of a predominant ancestry, we can simply keep individuals who are close enough to the overall center of all individuals (Figure \ref{fig:hist3}).
When doing so, we can cluster 91\% of the data into one cluster composed of 421,871 British, 12,039 Irish, 8351 ``Other White'', 1814 individuals of unknown ancestry, 467 ``White'' and 41 individuals of other self-reported ancestries.
This is made possible because we use the geometric median which is robust to outliers.

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{map-unmatched}}
	\caption{
		Percentage of individuals from the UK Biobank that could not been matched to any of the 26 1000G populations using our method, per country of birth (Field 20115). Countries in grey contain less than 30 individuals, therefore their percentages are not represented.
		Red points represent the locations of the 1000G populations, accessed from \url{https://www.internationalgenome.org/data-portal/population}. 
		Note that ``Gujarati Indian from Houston, Texas'' were manually moved to Gujarat (22.309425, 72.136230), ``Sri Lankan Tamil from the UK'' to Sri Lanka (6.927079, 79.861244), and ``Indian Telugu from the UK'' to (16.5, 79.5) to better reflect the location of their ancestors.
		Also note that ``Utah Residents with Northern and Western European Ancestry'', ``Americans of African Ancestry in SW USA'', ``African Caribbeans in Barbados'' and ``Mexican Ancestry from Los Angeles USA'' are probably not located at their ancestral location. \label{fig:map}}
\end{figure}

@@

Authors thank Alex Diaz-Papkovich, Clive Hoggart and others for their useful feedback.
This research has been conducted using the UK Biobank Resource under Application Number 41181.

@@

@@

\begin{figure}[h]
	\centerline{\includegraphics[width=0.95\textwidth]{compare-Euclidean-to-Fst2}}
	\caption{Comparing $F_{ST}$ to the squared Euclidean distance on the PCA space between centers of pairs of the 26 1000G populations. Distances are computed using a different number of Principal Components (PCs) \label{fig:compare-distK}}
\end{figure}

@@

%\begin{figure}[h]
%\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-GBR}}
%\caption{Histogram of squared distances (on a log scale) from the projected UK Biobank individuals to the geometric median of the GBR individuals from the 1000G data. The threshold in red assumes an $F_{ST}$ of 0.002. \label{fig:hist1}}
%\end{figure}
%
%\begin{figure}[h]
%	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-British}}
%	\caption{Histogram of squared distances (on a log scale) from the UK Biobank PC scores to the geometric median of the UKBB British individuals. The threshold in red assumes an $F_{ST}$ of 0.002. \label{fig:hist2}}
%\end{figure}

\begin{figure}[h]
	\centerline{\includegraphics[width=0.8\textwidth]{hist-dist-overall-center}}
	\caption{Histogram of (log) squared distances from the UK Biobank PC scores to the geometric median of the all UKBB individuals. Here we use a threshold at 7, based on visual inspection. Alternatively, a more stringent threshold at 6 could also be used. \label{fig:hist3}}
\end{figure}
